---
title: "Class 13: Genome Informatics"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Overview
The purpose of this lab session is to introduce a set of tools used in highthroughput sequencing and the process of investigating interesting gene variance in Genomics, using high throughput sequencing. The tool we will be utilizing is Galaxy.

We will be analyzing Childhood asthma, as several gene variants have been implicated in its pathogenesis.

## Identify genetic variants of interest 

The four candidate SNPs are : **rs12936231**, **rs8067378**, **rs9303277**, and **rs7216389** which were found by referencing Verlaan et al. (2009) and [OMIM](http://www.omim.org). 


THrough the ENSEMBLE page, these variants were found to overlap with or effect the function of these three genes : **ZPBP2**, **GSDMB**, and **ORMDL3**.

In the ENSEMBLE genome browser under the locations tab, the location of rs8067378 was found to be on *Chromosome 17: 39895095*; and that there are *(forward strand) A/G variants (43% G) different alleles* for rs8067378.

Under *sample genotypes* > Mexican Ancestry in Los Angeles, California it was discovered that 14% of the Mexican Ancestry in Los Angeles sample population (MXL) are homozygous for the asthma associated SNP (G|G).

```{r}
x <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
(sum(x[,2] == "G|G") / nrow(x)) * 100
```

##Initial RNA-seq data


**work in progress**
















