---
title: "Class 15: Pathway Analysis from RNA-Seq Results"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
#BiocManager::install("DESeq2")
library(DESeq2)
```

```{r}
#my first inclination:
#meta <- read.csv("GSE37704_metadata.csv")

#featcount <- read.csv("GSE37704_featurecounts.csv")
```

```{r}
#That of the project
metaFile <- "GSE37704_metadata.csv"
countFile <- "GSE37704_featurecounts.csv"

# Import metadata and take a peak
colData = read.csv(metaFile, row.names=1)
countData = read.csv(countFile, row.names =1)
head(colData)
```
We need the first column of the meta to match that of the count data; thus, there's an interferring 'length' column in the count data set that must be removed.

###Removing the length column
```{r}
countData <- as.matrix(countData[,-1])
head(countData)
```
An Additional way
```{r}
#countData$length <- NULL
#head(countData)
```

```{r}
zero.inds <- rowSums(countData) == 0

head(countData[-zero.inds, ])
```

```{r}
dds = DESeqDataSetFromMatrix(countData=countData,
                             colData=colData,
                             design=~condition)
dds = DESeq(dds)
```
HoxA1 knockdown versus control siRNA


```{r}
res = results(dds)

summary(res)
```
**Volcano plot**
```{r}
plot( res$log2FoldChange, -log(res$padj) )
abline(v=c(-2,2), col="green", lwd=2)
abline(h=-log(0.01), col="green", lwd=2)
```
Now coloring this plot: below the horizontal line = red outside of the vertical lilnes = blue

```{r}
#creating the custom color vector
mycols <- rep("gray", nrow(res) )

#red points for outside the cental chunk
mycols[ abs(res$log2FoldChange) > 2 ] <- "red"

# Blue points
inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

plot( res$log2FoldChange,  -log(res$padj), 
 col=mycols, ylab="-Log(P-value)", xlab="Log2(FoldChange)" )
```

###Adding Gene Annotation
```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
```


```{r}
#pulling up the 'cheat sheet'
columns(org.Hs.eg.db)
```

```{r}
res$symbol = mapIds(org.Hs.eg.db,
                    keys=row.names(res), # Our genenames
                     keytype="ENSEMBL",        # The format of our genenames
                     column="SYMBOL",          # The new format we want to add
                    multiVals="first")
```

```{r}
res$entrez = mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="ENTREZID",
                    multiVals="first")
```

```{r}
res$name =   mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="GENENAME",
                    multiVals="first")
```

```{r}
head(res, 10)
```

```{r}
res = res[order(res$pvalue),]
write.csv(res, file="deseq_results.csv")
```

##Kegg Pathway
```{r}
library(pathview)
```

```{r}
library(gage)
library(gageData)

data(kegg.sets.hs) #KEGG data set we're using
data(sigmet.idx.hs) #flies data set we're using

# Focus on signaling and metabolic pathways only
kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]

# Examine the first 3 pathways
head(kegg.sets.hs, 3)
```
See how there are more genes involved in Purine metabolism compared to Caffeine metabolism?

Works with ENtrezIDs

The gage package wants a vector as input with our numeric scores of interest.
```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)

# Get the results
keggres = gage(foldchanges, gsets=kegg.sets.hs)

attributes(keggres)
```

```{r}
head(keggres$less)
```

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04110")
```


```{r}
# A different PDF based output of the same data
pathview(gene.data=foldchanges, pathway.id="hsa04110", kegg.native=FALSE)
```

```{r}
# Focus on top 5 upregulated pathways here for demo purposes only
keggrespathways <- rownames(keggres$greater)[1:5]

# Extract the 8 character long IDs part of each string

keggresids = substr(keggrespathways, start=1, stop=8) #Extract or replace substrings in a character vector.
keggresids
```

```{r}
# pass these IDs in keggresids to the pathview() function to draw plots for all the top 5 pathways
pathview(gene.data=foldchanges, pathway.id=keggresids, species="hsa")
```
##Section 3. Gene Ontology
Doing a similar task but with a different resource
```{r}
data(go.sets.hs)
data(go.subs.hs)

# Focus on Biological Process subset of GO
gobpsets = go.sets.hs[go.subs.hs$BP]

gobpres = gage(foldchanges, gsets=gobpsets, same.dir=TRUE)

lapply(gobpres, head)
```
##Section 4. Reactome Analysis
Reactome is database consisting of biological molecules and their relation to pathways and processes. Reactome, such as many other tools, has both online software and R packages available.
```{r}
#output the list of significant genes at the 0.05 level as a plain text file
sig_genes <- res[res$padj <= 0.05 & !is.na(res$padj), "symbol"]
print(paste("Total number of significant genes:", length(sig_genes)))
```

```{r}
write.table(sig_genes, file="significant_genes.txt", row.names=FALSE, col.names=FALSE, quote=FALSE)
```
to perform pathway analysis online go to the Reactome website (https://reactome.org/PathwayBrowser/#TOOL=AT)
