---
title: "Class 11: Foundatons of Bioinformatics"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Introduction to the RCSB Protein Data Bank (PDB)
**Visualizing the HIV-1 protease structure**

First, let's load up the packages we'll be utilizing
```{r}
library(bio3d)
library(bio3d.view)
```

as well as accessing our pdb file
```{r}
pdb <- read.pdb("1hsg")

pdb
```

### Familizarizing ourselves with accessing this data

How to find the attributes of this object
```{r}
attributes(pdb)
```

Accessing individual attribtures
```{r}
head(pdb$atom)
```

Print a summary of the coordinate data in $xyz
```{r}
pdb$xyz
```

```{r}
# Examine the row and column dimensions
dim(pdb$xyz)

# Print coordinates for the first two atom
pdb$xyz[ 1, atom2xyz(1:2) ]
```

###Using Atom Select
```{r}
# Select all C-alpha atoms (return their indices)
ca.inds <- atom.select(pdb, "calpha")
ca.inds
```

```{r}
# Print details of the first few selected atoms
head( pdb$atom[ca.inds$atom, ] )

# And selected xyz coordinates
head( pdb$xyz[, ca.inds$xyz] )
```

**Fine-grained
```{r}
# Select chain A
a.inds <- atom.select(pdb, chain="A")
# Select C-alphas of chain A
ca.inds <- atom.select(pdb, "calpha", chain="A")
# We can combine multiple selection criteria to return their intersection
cab.inds <- atom.select(pdb, elety=c("CA","CB"), chain="A",
resno=10:20)
```

###Creating a protein- and ligand-only PDB file through R
We will use the Bio3D function *atom.select()* followed by the write.pdb() functions, which will isolate the components we want and create a new file for us to access later.
```{r}
prot <- atom.select(pdb, "protein", value=TRUE)
prot #protein only

write.pdb(prot, file="1hsg_protein.pdb")


lig <- atom.select(pdb, "ligand", value=TRUE)
lig #ligand only

write.pdb(lig, file="1hsg_ligand.pdb")
```

Let's take a look at the files we generated
```{r}
view(lig)

view(prot)

view(pdb, "overview", col="sse")
```

```{r}
# Load the package
pdb <- read.pdb("1hel")
# Normal mode analysis calculation
modes <- nma(pdb)
m7 <- mktrj(modes,
 mode=7,
 file="mode_7.pdb")
view(m7, col=vec2color( rmsf(m7) ))
```

###Working with multiple PDB files
**aligning multiple structures**
```{r}
# Download some example PDB files
ids <- c("1TND_B","1AGR_A","1TAG_A","1GG2_A","1KJY_A","4G5Q_A")
files <- get.pdb(ids, split = TRUE)

# Extract and align the chains we are interested in
pdbs <- pdbaln(files, fit = TRUE) #this superimposes the molecules

# Print to screen the superimposion of the 'pdbs' object
pdbs
```

Let's take a peak at the new data, then go on to analyze it more deeply. Ultimately, we want to produce a PCA to analyze the most significant structural variations within this superimposition of structures we have generated.
```{r}
# Access the first 5 rows, and 8 columns
pdbs$ali[1:5, 1:8]
```

```{r}
# Associated residues numbers
pdbs$resno[1:5, 1:8]
```

###Basic structure Analysis
```{r}
# Calculate sequence identity
seqidentity(pdbs)
```

```{r}
# Calculate RMSD
rmsd(pdbs)
```

```{r}
# Calculate RMSD
rd <- rmsd(pdbs)
# Clustering
hc <- hclust(as.dist(rd))
grps <- cutree(hc, k=3)
# Plot results as dendrogram
hclustplot(hc, k=3)
```

**Principal component analysis**
```{r}
# Perform PCA
pc.xray <- pca(pdbs)

# Plot our results summary (PCA score plot and scree-plot)
plot(pc.xray)
```

```{r}
# Visualize first principal component
pc1 <- mktrj(pc.xray, pc=1, file="pc_1.pdb")
```

```{r}
library(bio3d.view)
# Structural displacements captured by PC1
view(pc1)
# The rglwidget() function from the rgl
# package will show output in your Rmd
# notebook and rendered html_output
# documents
#library(rgl)
#rglwidget(pc1)
```
